\chapter{Conclusion}
\section{Summary of Findings}
In this dissertation, we have biologically motivated the need to consider time delays within cellular gene-expression processes. We have also mathematically shown the effects these time delays can have on Turing pattern formation, which is especially significant when considering patterning events that occur on a fast timescale.

Using Turing instability analysis for the LI variant of the Schnakenberg model, we have shown that an increasing fixed time delay causes an expansion of the Turing space, resulting in a larger parameter set capable of exhibiting Turing instabilities. In contrast, we have shown that the same increasing time delay results in an increase in the time taken until onset of patterning. It was shown through linear theory and full numerical solutions that this relationship between increasing time delay and onset of patterning is linear.

Our study also shows that, through a systematic exploration using full numerical solutions, that a variation in initial conditions, boundary conditions or history function does effect the \textit{type} of pattern formation seen. However, the relationship between onset of patterning and fixed time delay seems to be robust to these variations, and thus they do not provide a means of remedying an increase in time-to-pattern as a result of time delay.

In introducing both a symmetric and skewed truncated Gaussian distribution, we applied novel linear analysis to show that the distribution used does not matter. Namely, the results with a time delay modelled as a distribution do not significantly vary from those produced with a fixed time delay.

Finally, by using the Gierer-Meinhardt model, we demonstrated the importance of the position of time-delayed terms within a reaction-diffusion mechanism on the Turing space. An increasing time-delay was shown to both expand and shrink the Turing space of two differnt variants of the GM model, and it was shown that the shrinking or expansion of the parameter space was dependent on the spatially homogeneous model only.

\section{Future Work}
